---
title: "Functional enrichment practical"
author: "Olivier Sand and Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    self_contained: yes
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: "hide"
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../slides.css
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  powerpoint_presentation:
    slide_level: 2
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: DUBii 2021
font-family: Garamond
transition: linear
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE, echo=FALSE, eval=TRUE}

options(width = 300)
# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/enrich-',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  results = TRUE, 
  comment = "")

options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

```


```{r libraries, echo=FALSE, eval=TRUE}

#### Load libraries ####
message("Loading libraries")
requiredLib <- c("knitr","gprofiler2")
for (lib in requiredLib) {
  if (!require(lib, character.only = TRUE)) {
    install.packages(lib, )
  }
  require(lib, character.only = TRUE)
}

library(gprofiler2)

```

## Goal of this practical

This practical aims at showing how to analyze functional enrichment on a list of genes with `gProfiler`.

## Study case : The Cancer Gene Atlas (TCGA) Breast Invasive Cancer (BIC) data

Description : <https://du-bii.github.io/study-cases/Homo_sapiens/TCGA_study-case/>

## References

- R package : <https://cran.r-project.org/web/packages/gprofiler2>

- Reference manual: <https://cran.r-project.org/web/packages/gprofiler2/gprofiler2.pdf>

- Vignette: <https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html>

## Load (clusters of differentially) expressed genes

First, you need to load the gene list, which can be found on the github repository, in the following file. 

- 1000 top DEG genes: <https://raw.githubusercontent.com/DU-Bii/module-3-Stat-R/master/stat-R_2021/data/TCGA_BIC_subset/BIC_edgeR_DEG_top_1000_geneIDs.txt>


```{r data_loading}

#### Load the gene list ####
message("Loading gene list")
## URL of the data folder on github
data_folder_url <- 'https://raw.githubusercontent.com/DU-Bii/module-3-Stat-R/master/stat-R_2021/data/TCGA_BIC_subset/'

gene_list_file <- "BIC_edgeR_DEG_top_1000_geneIDs.txt"

## Trick : the file is read as a one-column table, and we transform it to a vector with unlist()
gene_list <- unlist(read.delim(file = file.path(data_folder_url, gene_list_file), header = FALSE))

## Check the length of the gene list
length(gene_list)

```


## Gene list functional enrichment analysis with gost

1. Use the `gost()` function to get the enrichment results.

- Note: we use FDR as multiple testing correction.

```{r gost}

#### Run enrichment analysis with gost() ####
message("Running the enrichment analysis")
## Run the analysis
gostres <- gost(query = gene_list, 
                organism = "hsapiens", 
                ordered_query = FALSE, 
                multi_query = FALSE, 
                significant = TRUE, 
                exclude_iea = FALSE, 
                measure_underrepresentation = FALSE, 
                evcodes = FALSE, 
                user_threshold = 0.05, 
                correction_method = "fdr", 
                domain_scope = "annotated", 
                custom_bg = NULL, 
                numeric_ns = "", 
                sources = NULL, 
                as_short_link = FALSE)

## Check the structure of the result
names(gostres)
names(gostres$result)
kable(summary(gostres))
## Check the most significant results, formating for kable
#head(gostres$result)
kable(head(gostres$result), digits = c(0, 0, 15, 7, 7, 7, 3, 3, 0, 0, 7, 7, 0, 0))
# kable(head(gostres$result), 
#       format = c("%s", "%s", "%e", "%d", "%d", "%d", "%.3f", "%.3f", "%s", "%s", "%d", "%d", "%s", "%s"))
names(gostres$meta)

```

2. Visualize the results as an interactive plot


```{r gostplot, fig.width=8, fig.height=5, out.width="100%", fig.cap="Dynamic figure of the enrichment analysis. "}

### Visualization ###
message("Generating a dynamic plot of the results")
# Plot the results
gostplot(gostres, capped = TRUE, 
         interactive = TRUE)

```


We can now tune the parameters to get some additional information about the evidence code and the list of genes at the intersection between the query set and the functional classes. Note that this generates a table with very large columns. 

```{r gostplot_intersect}

### With the input gene IDs that intersect with the corresponding functional term ###
message("Running the enrichment analysis showing the intersections")
# With intersection (can be time consuming)
gostres2 <- gost(query = gene_list, 
                organism = "hsapiens", 
                ordered_query = FALSE, 
                multi_query = FALSE, 
                significant = TRUE, 
                exclude_iea = FALSE, 
                measure_underrepresentation = FALSE, 
                evcodes = TRUE, 
                user_threshold = 0.05, 
                correction_method = "fdr", 
                domain_scope = "annotated", 
                custom_bg = NULL, 
                numeric_ns = "", sources = NULL)

kable(head(gostres2$result), digits = c(0, 0, 15, 7, 7, 7, 3, 3, 0, 0, 7, 7, 0, 0, 0, 0))

```

3. Get a web link to share the results with your coworkers


```{r web_link}

### Web link ###
message("Getting a web link to view the results")
# Get a web link
gostres_link <- gost(query = gene_list, 
                      organism = "hsapiens", 
                      ordered_query = FALSE, 
                      multi_query = FALSE, 
                      significant = TRUE, 
                      exclude_iea = FALSE, 
                      measure_underrepresentation = FALSE, 
                      evcodes = TRUE, 
                      user_threshold = 0.05, 
                      correction_method = "fdr", 
                      domain_scope = "annotated", 
                      custom_bg = NULL, 
                      numeric_ns = "", 
                      sources = NULL,
                      as_short_link = TRUE)

gostres_link

```

4. Make a static plot 

```{r static_plot, fig.width=8, fig.height=5, out.width="100%", fig.cap="Static figure of the enrichment analysis. "}

### Static plot ###
message("Generating a static plot")
# Static ggplot object
p <- gostplot(gostres, capped = FALSE, interactive = FALSE)
p

```

5. Highlight the `GO:0071840` and `HPA:0300000` terms

```{r highlight_terms, fig.width=8, fig.height=5, out.width="100%", fig.cap="Static figure of the enrichment analysis with highlighted terms. "}
### Highlight terms ###
message("Highlighting terms")
# Highlight terms
pp <- publish_gostplot(p, 
                       highlight_terms = c("GO:0071840", "HPA:0300000"), 
                       width = NA, 
                       height = NA, 
                       filename = NULL )
#pp
```

6. Produce a custom table with choosen results

```{r custom_table, fig.width=10, fig.height=5, out.width="100%", fig.cap="Custom table of the enrichment analysis with choosen results. "}
# Table
publish_gosttable(gostres, 
                  highlight_terms = gostres$result[c(1:2,10,50:52,64,69,71),],
                  use_colors = TRUE, 
                  show_columns = c("source", "term_name", "term_size", "intersection_size"),
                  filename = NULL)

```

If you feel adventurous, you can try to do the same with multiple gene lists (for instance, clusters from the previous practical).

## Save your session info

For the sake of traceability, store the specifications of your R environment in the report, with the command `sessionInfo()`. This will indicate the version of R as well as of all the libraries used in this notebook. 

```{r session_info}

sessionInfo()

```
