---
title: "Differential analysis of proteomics data"
author: "Jacques van Helden"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    code_folding: hide
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: no
    code_folding: hide
    css: ../../slides.css
  slidy_presentation:
    font_adjustment: 0   ## set to negative/positive values for smaller/bigger fonts
    duration: 45
    self_contained: no
    code_folding: hide
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  ioslides_presentation:
    self_contained: no
    code_folding: hide
    css: ../../slides.css
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: DUBii -- Statistics with R
font-family: Garamond
transition: linear
---

```{r include=FALSE, echo=FALSE, eval=TRUE}

## Install required packages
packages <- c("knitr", "vioplot")
for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    message("Installing missing package: ", pkg)
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

## Set options for knitr
options(width = 300)
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  out.width = "80%",
  fig.align = "center", 
  fig.path = "figures/data-structures_",
  size = "tiny", 
  echo = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  results = TRUE, 
  comment = "")
# knitr::asis_output("\\footnotesize")
```

## Data sources

Study case: 

- Stare, T., Ramšak, Ž., Križnik, M. et al. Multiomics analysis of tolerant interaction of potato with potato virus Y. Sci Data 6, 250 (2019). https://doi.org/10.1038/s41597-019-0216-1 (<https://www.nature.com/articles/s41597-019-0216-1>)

- The metadata and data area available on the PRIDE database (<https://www.ebi.ac.uk/pride/archive/projects/PXD015221>).

- The data was downloaded from the FTP site (<ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2019/09/PXD015221>). 




## Data loading

### Data location

```{r}
## Local dir containing the data
dataDir <- "data"
# list.files(dataDir)

## Data file (table of protein quantifications)
data.file <- "ProteomeDiscoverer_Multiconsensus_from_18_Reports_PEP_INFO_protein-table.csv"

## Metadata file (sample descriptions)
metadata.file <- "Metadata_sample_names.csv" 

```

### Load and check the metadata

```{r}
## Load metadata table
metadata <- read.csv(file = file.path(dataDir, metadata.file))
# View(metadata)
# names(metadata)

# Select the relevant columns for our analysis
fields <- c("proteomicsPool", 
            "Genotype",
            "Treatment", 
            "BiosampleName")
poolDescriptions <- unique(metadata[, fields])
# dim(poolDescriptions)
# View(poolDescriptions)

poolNames <- as.vector(poolDescriptions$proteomicsPool)

## Define pool condition as the combination of genotype + treatment
poolDescriptions$condition <- paste(
  sep = "_", 
  poolDescriptions$Genotype,
  poolDescriptions$Treatment)

## Suppress the leadin "D-", which is found everywhere and this not informative
poolDescriptions$condition <- sub(pattern = "^D-", replacement = "", x = poolDescriptions$condition)

## Build informative labels for the pools
poolLabels <- paste(sep = "_", 
                    poolDescriptions$condition,
                    poolDescriptions$proteomicsPool)
poolLabels <- sub(pattern = "_pool", replacement = "_", x = poolLabels)
names(poolLabels) <- poolNames # use poolNames as indices

## Assign the pool labels as rownames for the metadata
rownames(poolDescriptions) <- poolLabels

## Define condition-specific colors
conditions <- sort(unique(poolDescriptions$condition))
conditionColors <- rainbow(n = length(conditions))
names(conditionColors) <- conditions

## Assign a color to each pool according to its condition
poolDescriptions$color <- conditionColors[poolDescriptions$condition]

```

The metadata table contains `r nrow(metadata)` rows, each describing one sample. However, samples were pooled by pairs for the proteomics analysis, so that the actual data contains `r nrow(poolDescriptions)` pools. The table below provides the description of the pools. 

```{r}
kable(poolDescriptions, caption = "Description of the sample pools.")
```

The data contains 3 replicates per condition, where a condition is a combination of a genotype (COI, NAHG or WT) and a treatment (MOCK or PVY).

```{r}
library(knitr)
kable(table(poolDescriptions[, c("Genotype", "Treatment")]), 
      caption = "Number of replicates per condition. Columns: treatment (MOCK or virus infection). Rows: genotype. ")
```

### Loading the data table


```{r}
data <- read.csv(file = file.path(dataDir, data.file))
# names(data)
# head(data)

## Extract the quantities in a separat data table to facilitate the statistical analysis
proteinQuant <- data[, poolNames]
rownames(proteinQuant) <- data[, "Accession"]
# dim(proteinQuant)
# View(proteinQuant)

## Replace pool names by pool labels in the protein table
names(proteinQuant) <- poolLabels[names(proteinQuant)]
# View(proteinQuant)



```

The data table contains some descriptive columns, followed by  `r nrow(proteinQuant)` columns providing the quantities of `r nrow(proteinQuant)` proteins (rows) in each sample pool (columns). 


## Distributions of quantities per sample

Draw a box plot of the protein quantities using `boxplot()` with its the default options. Then read the help page (`help(boxplot)`) and fine-tune the options to get a more informative result.  

```{r boxplot_ugly, fig.width=10, fig.height=5, out.width="80%", fig.cap="Boxplot showing the distribution of counts per sample for all the detected proteins. The result is not readable, we will do better in the next fiugure. "}
boxplot(proteinQuant)
```

```{r boxplot, fig.width=10, fig.height=6, out.width="100%", fig.cap="Boxplot showing the distribution of counts per sample for all the detected proteins"}

## Store initial parameter values
initialParameters <- par(no.readonly = TRUE)
# par("mar")

## Modify the parameters
par(mar = c(4.1, 7, 1, 1))
par(mfrow = c(1, 2))
boxplot(proteinQuant, 
        col = poolDescriptions$color,
        xlab = "Raw counts",
        horizontal = TRUE, 
        cex.axis = 0.8,
        las = TRUE)
boxplot(log2(proteinQuant), 
        col = poolDescriptions$color,
        xlab = "log2(counts)",
        horizontal = TRUE, 
        cex.axis = 0.8,
        las = TRUE)
par(initialParameters)
```


```{r vioplot, fig.width=10, fig.height=6, out.width="100%", fig.cap="Violin showing the distribution of counts per sample for all the detected proteins"}

## Store initial parameter values
# par("mar")

## Modify the parameters
par(mar = c(4.1, 7, 1, 1))
par(mfrow = c(1, 2))
vioplot(proteinQuant, 
        col = poolDescriptions$color,
        xlab = "Raw counts",
        horizontal = TRUE, 
        cex.axis = 0.8,
        las = TRUE)
vioplot(log2(proteinQuant), 
        col = poolDescriptions$color,
        xlab = "log2(counts)",
        horizontal = TRUE, 
        cex.axis = 0.8,
        las = TRUE)

par(initialParameters)
```

## Normalisation

### Mean-based normalisation

### Median-based normalisation

```{r median_norm}

## Compute the median per pool
medianPerPool <- apply(X = proteinQuant, 2, median, na.rm = TRUE)
median <- median(unlist(proteinQuant), na.rm = TRUE)

proteinQuantMedianNorm <- data.frame(t(t(proteinQuant) * median / medianPerPool))
# dim(proteinQuantMedianNorm)
# summary(proteinQuantMedianNorm)
# apply(proteinQuantMedianNorm, 2, median, na.rm = TRUE)

## Modify the parameters
par(mar = c(4.1, 7, 4.1, 1))
par(mfrow = c(1, 2))
class(proteinQuant)
dim(proteinQuant)
vioplot(proteinQuantMedianNorm, 
        main = "Median-based scaling",
        col = poolDescriptions$color,
        xlab = "Raw counts",
        horizontal = TRUE, 
        cex.axis = 0.8,
        las = TRUE)
vioplot(log2(proteinQuantMedianNorm), 
        main = "Median-based scaling",
        col = poolDescriptions$color,
        xlab = "log2(counts)",
        horizontal = TRUE, 
        cex.axis = 0.8,
        las = TRUE)

par(initialParameters)

```



## Multidimensional scaling with PCA

A first exploratory approach is to project the samples on a 2-dimensional plane, where the distance between samples approximately corresponds to the differences between their proteomics profiles. 


