---
title: "Tutorial : correlation and regression"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    self_contained: no
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: "hide"
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../slides.css
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  powerpoint_presentation:
    slide_level: 2
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: DUBii 2020
font-family: Garamond
transition: linear
editor_options: 
  chunk_output_type: console
---




```{r settings, include=FALSE, echo=FALSE, eval=TRUE}

options(width = 300)
# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/pavkovic2019_',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  results = TRUE, 
  comment = "")

options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

```



```{r libraries, echo=FALSE, eval=TRUE}
requiredLib <- c("knitr")
for (lib in requiredLib) {
  if (!require(lib, character.only = TRUE)) {
    install.packages(lib, )
  }
  require(lib, character.only = TRUE)
}

```


## Goals of this tutorial

This tutorial provides a quick reminder of the basic concepts of correlation and regression analysis, illustrated by an empirical study of a multi-omics dataset. 


## Study case : mouse kidney 

As study case for this tutorial, we will use multi-omics data from a study published by [Pavkovic et al. (2019)](https://doi.org/10.1038/s41597-019-0095-5), which combines transctiptomics (RNA-seq) and proteomics approaches to understand the molecular mechanisms underlying the kidney fibrosis pathology. 

The authors applied two commonly used treatments to induce kidney fibrosis in mouse: 

- a reversible chemical-induced injury model, denoted as **FA** for folic acid induced nephropathy;  

- an irreversible surgically-induced fibrosis model, denoted as **UUO** for unilateral ureteral obstruction.

## References and data sources


-  **Reference**: Pavkovic, M., Pantano, L., Gerlach, C.V. et al. Multi omics analysis of fibrotic kidneys in two mouse models. Sci Data 6, 92 (2019) <https://doi.org/10.1038/s41597-019-0095-5>

- **Mouse fibrotic kidney browser:** <http://hbcreports.med.harvard.edu/fmm/>

- Data on Zenodo: <https://zenodo.org/record/2592516>

## Data preparation

A description of the study case can be found in the *[Mus musculus section](https://du-bii.github.io/study-cases/#mus-musculus)* of the the [DUBii study cases repository](https://du-bii.github.io/study-cases/) repository. 

We also provide there a detailed explanation of the [data preparation steps](https://du-bii.github.io/study-cases/Mus_musculus/pavkovic_2019/Rmd/prepare-data_pavkovic_2019.html): 

- downloading the data from its original source repository, 
- exploring the datasets it with various graphical representtions, 
- computing some descriptive statistics on the different samples, 
- pre-processing,
- storing the results in a memory image. 

## Approach for this tutorial

This tutorial will consist of exercises that can be realised in a stepwise way, with alternance of working sessions and live demos of the solutions, in order to make sure that all the trainees acquire each step.

## Data exploration

Before computing any descriptive parameter on a dataset, I generallyi attempt to get a picture of the whole distribution.


### Exercise : data loading

```{r data_loading}

```


### Histogram


### Box plot


### Scatter plot





