---
title: "R markdown report template"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    self_contained: no
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: "hide"
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../slides.css
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  powerpoint_presentation:
    slide_level: 2
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: DUBii 2020
font-family: Garamond
transition: linear
editor_options: 
  chunk_output_type: console
---



```{r settings, include=FALSE, echo=FALSE, eval=TRUE}

options(width = 300)
# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/pavkovic2019_',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  results = TRUE, 
  comment = "")

options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

```



```{r libraries, echo=FALSE, eval=TRUE}
requiredLib <- c("knitr", "stringr", "vioplot")
for (lib in requiredLib) {
  if (!require(lib, character.only = TRUE)) {
    install.packages(lib, )
  }
  require(lib, character.only = TRUE)
}

```

## Introduction

### Nice features

- The header contains reasonable parameters for different types of compilations with knitr (pdf report, html report, slidy presentation, ...)
- automatic update of the compilation date
- floating table of contents for HTML slides
- code folding (collapsed by default) enables to see the code used to generate the report, whilst preserving a nice HTML report (does not work for pdf or word exports)


## Could be improved

- export of pdf slides
- slides with two columns (one for the text and one for figures

### Tuning

- self_contained: yes makes a self-contained HTML, whcih is heavier (>2Mb even for very small report) but can be sent by email



