---
title: "Start R -- First steps with R and RStudio"
author: "Claire Vandiedonck & Jacques van Helden"
date: "`r Sys.Date()`"
output:
  slidy_presentation:
    self_contained: no
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../slides.css
  html_document:
    self_contained: no
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    widescreen: yes
  ioslides_presentation:
    self_contained: no
    css: slides.css
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: DUBii -- Statistics with R
font-family: Garamond
transition: linear
---

```{r include=FALSE, echo=FALSE, eval=TRUE}
library(knitr)

options(width = 300)
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  out.width = "80%",
  fig.align = "center", 
  fig.path = "figures/start-R_",
  size = "tiny", 
  echo = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  results = TRUE, 
  comment = "")
# knitr::asis_output("\\footnotesize")

# dir.main <- "~/stat1"
# dir.slides <- file.path(dir.main, "slides")
# setwd(dir.slides)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Connection to IFB RStudio server

In a Web browser, open a connection to the IFB RStudio server

<https://rstudio.cluster.france-bioinformatique.fr/>

This requires your login and password. 


## Navigating in your Unix account


```{r eval=FALSE}
## Where am I now?
getwd()

## Where is my home?
Sys.getenv("HOME")

## Who am I?
Sys.getenv("LOGNAME")

## Go home
setwd(Sys.getenv("HOME"))
setwd("~")  ## An easier way

```

### Explanations

- `Sys.getenv()` returns all the environment variables
- `$HOME` is a Unix variable environment indicating the home directory
- `$LOGNAME` is my login name (did you guess?)
- in Unix, `~` denotes your home directory (equivalent to `$HOME`)


## Creating a work space for this course

We will create specific directory to store the data and results of the practicals for this course.

```{r eval=FALSE}
## Define the directory for this course
courseDir <- file.path("~", "dubii20", "stat-R", "session1")
print(courseDir) ## Check the result

## Create the directory
dir.create(courseDir) ## THIS WILL LIKELY NOT WORK
dir.create(path = courseDir, 
           recursive = TRUE, ## Create the parent directories
           showWarnings = FALSE) ## Don't shout it already exists

## Go to this directory
setwd(courseDir)
getwd() ## Check that you are in the right place

```

