---
title: "Tutorial: exploration of multi-omics data"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    self_contained: no
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: "hide"
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../slides.css
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  powerpoint_presentation:
    slide_level: 2
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: DUBii 2020
font-family: Garamond
transition: linear
editor_options: 
  chunk_output_type: console
---




```{r settings, include=FALSE, echo=FALSE, eval=TRUE}

options(width = 300)
# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/pavkovic2019_',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  results = TRUE, 
  comment = "")

options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

```



```{r libraries, echo=FALSE, eval=TRUE}
requiredLib <- c("knitr")
for (lib in requiredLib) {
  if (!require(lib, character.only = TRUE)) {
    install.packages(lib, )
  }
  require(lib, character.only = TRUE)
}

```


## Goals of this tutorial

This tutorial provides a quick reminder of the basic concepts of correlation and regression analysis, illustrated by an empirical study of a multi-omics dataset. 


## Study case : mouse kidney 

As study case for this tutorial, we will use multi-omics data from a study published by [Pavkovic et al. (2019)](https://doi.org/10.1038/s41597-019-0095-5), which combines transctiptomics (RNA-seq) and proteomics approaches to understand the molecular mechanisms underlying the kidney fibrosis pathology. 

The authors applied two commonly used treatments to induce kidney fibrosis in mouse: 

- a reversible chemical-induced injury model, denoted as **FA** for **folic acid** induced nephropathy;  

- an irreversible surgically-induced fibrosis model, denoted as **UUO** for **unilateral uretral obstruction**.

## References and data sources


-  **Reference**: Pavkovic, M., Pantano, L., Gerlach, C.V. et al. Multi omics analysis of fibrotic kidneys in two mouse models. Sci Data 6, 92 (2019) <https://doi.org/10.1038/s41597-019-0095-5>

- **Mouse fibrotic kidney browser:** <http://hbcreports.med.harvard.edu/fmm/>

- Data on Zenodo: <https://zenodo.org/record/2592516>

## Data preparation

A description of the study case can be found in the *[Mus musculus section](https://du-bii.github.io/study-cases/#mus-musculus)* of the the [DUBii study cases repository](https://du-bii.github.io/study-cases/) repository. 

We also provide there a detailed explanation of the [data preparation steps](https://du-bii.github.io/study-cases/Mus_musculus/pavkovic_2019/Rmd/prepare-data_pavkovic_2019.html): 

- downloading the data from its original source repository, 
- exploring the datasets it with various graphical representtions, 
- computing some descriptive statistics on the different samples, 
- pre-processing,
- storing the results in a memory image. 


## Data file naming

We prepared the data from Pavkovic as a text file with tab-separated values (**tsv** files). 

All the files are available on github:
- <https://github.com/DU-Bii/module-3-Stat-R/tree/master/stat-R_2021/data/pavkovic_2019>

The files are named according to the following convention:

- the prefix indicate the data type

    - **fa**: folic acid induced nephropathy (reversible), transcriptome data
    - **pfa**: folic acid induced nephropathy, proteome data
    - **uuo**: unilateral uretral obstruction (irreversible), transcriptome data
    - **puuo**: unilateral uretral obstruction, proteome data

- The suffix indicates the data  normalisation

    - **raw**: transcriptome counts provided by the authors (note: not integer, because their raw data was already somehow normalized)
    - **normalized¨**: transcriptome counts standardized to achieve the same third quartile for each sample
    - **log2**: log2 transformation for proteome data
    
- the  **metadata** files contain a short description of each sample (one row per sample). Note that the last column contains a sample-specific color specification in [hexadecimal web color code](https://en.wikipedia.org/wiki/Web_colors#Hex_triplet) to facilitate the drawings. Don't hesitate to chose [other colors](https://htmlcolorcodes.com/) according to your personal taste. 

## Approach for this tutorial

This tutorial will consist of exercises that can be realised in a stepwise way, with alternance of working sessions and live demos of the solutions, in order to make sure that all the trainees acquire each step.

## Data exploration

Before computing any descriptive parameter on a dataset, I generallyi attempt to get a picture of the whole distribution.


### Finding a data file on github

We will provide here a magic recipe to download the data from the github repository to your local folder, and to load it in R. 


```{r data_loading}
## specify the base URL from which data files can be downloaded 
urlBase <- "https://github.com/DU-Bii/module-3-Stat-R/raw/master/stat-R_2021/data/pavkovic_2019"


## Choose a specific data file
dataPrefix <- "pfa" ## proteome data of folic-acid treated mouse
dataSuffix <- "model" ## no normalization
fileName <- paste0(dataPrefix, "_", dataSuffix, "_counts.tsv.gz")

## Compose the URL to download the file from github
url <- file.path(urlBase, fileName)

message("URL: ",  url)

```

### Loading a data file directly from github

Now we defined the URL, we can easily load the file directly from github to a data frame in our R environement. 

```{r load_from_github, eval=FALSE}
## this requires to load a specific package
if (!require("data.table")) {
  install.packages("data.table")
}
library(data.table)

pfa <- fread(url, header = TRUE, sep = "\t")

```



### Downloading a data file and storing it locally once forever

We can now download the data file to a local folder, but we would like to do this only once.

```{r download_only_once}
## Specify the path relative to your home directory (~)
localFolder <- "~/DUBii-m3_data/pavkovic_2019"
localFile <- file.path(localFolder, fileName)

## Create the local data folder if it does not exist
dir.create(localFolder, showWarnings = FALSE, recursive = TRUE)

## Download the file ONLY if it is not already there
if (!file.exists(localFile)) {
  message("Downloading file from github to local file\n\t", 
          localFile)
  download.file(url = url, destfile = localFile)
} else {
  message("Local file already exists, no need to download\n\t", 
          localFile)
}

```


### Loading the local copy of your data file

```{r load_local_file}
## Load the data from the local file
pfa <- read.delim(file = localFile, header = TRUE, sep = "\t")

```


### Exercise: writing functions to download and load a data file

Write two functions that will respectively download a file from a remote location to a local folder, but do this only if the local file is not yet present there. 

Use this function to load 
- the raw proteome data from folic acid treatment
- the proteome metadata


```{r exercise_download-load_function}
## Your turn to program

```


### Solution: implement functions to automagically download and load a data file

```{r solution_download-load_functions}
#' @title Download a file only if it is not yet here
#' @author Jacques van Helden
#' @param urlBase base of the URL, that will be prepended to the file name
#' @param fileNAme name of the file (should not contain any path)
#' @param localFolder path of a local folder where the file should be stored
#' @return the function returns the path of the local file, built from localFolder and fileName
downloadOnlyOnce <- function(urlBase, 
                             fileName,
                             localFolder) {

  ## Define the source URL  
  url <- file.path(urlBase, fileName)
  message("Source URL: ",  url)

  ## Define the local file
  localFile <- file.path(localFolder, fileName)
  
  ## Create the local data folder if it does not exist
  dir.create(localFolder, showWarnings = FALSE, recursive = TRUE)
  
  ## Download the file ONLY if it is not already there
  if (!file.exists(localFile)) {
    message("Downloading file from github to local file\n\t", 
            localFile)
    download.file(url = url, destfile = localFile)
  } else {
    message("Local file already exists, no need to download\n\t", 
            localFile)
  }
  
  return(localFile)
}
```

We can now use this function to download and load the different data files. 


## Graphical exploration of the data

### Histogram

Draw an histogram of the **puuo** data, all samples together. 

```{r puuo_hist_all-samples}

```



### Box plot


### Scatter plot


## Save your session info

As usual

```{r session_info}
sessionInfo()
```




